---
title: "Spotify Song Duration 60s vs Today"
author: "Layne Larson"
date: "2024-04-17"
output: 
pdf_document:
  latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




First lets start with the analysis on the 1960s song length. We will run through
a gibbs sampler using the data from the 1960s playlist, assess the convergence 
and mixing, and then plot the posterior density vs. the prior.
```{r}
library(invgamma)
set.seed(251)
top_60 <- read.csv("/Users/laynelarson12/Documents/Resumes and Stuff/Data Analysis Projects/Data Analysis Projects/all_out_60s.csv")
top_today <- read.csv("/Users/laynelarson12/Documents/Resumes and Stuff/Data Analysis Projects/Data Analysis Projects/todays_top_hits.csv")

# First we need to extract just the duration column as our data. Also lets 
# convert from milliseconds into seconds so the data is cleaner to work with

duration_60 <- top_60$Duration..ms./1000
duration_today <- top_today$Duration..ms./1000


# Prior parameter values
# song length is measured in milliseconds, and we believe songs in the 60s to be
# around 3 minutes in length. So lets choose a prior m of 180.
# then we would expect songs to vary by a minute on either end, so lets have a
# variance v of (120/6)^2 = 400
m_60 <- 180; v_60 <- 400
# for variance, lets use a=2 and b=3600 to approximate a mean variance of 3600,
# or a standard deviation for the sigma2 of 60.
a <- 2 ; b <- 3600

# use a gibbs sampler to find the posterior
unit8.gibbs <- function(data,m=0,v=1000,a=1,b=1,M=5000, burn=1000){
  ybar <- mean(data); n <- length(data)
  
  mu <- numeric()
  sigma2 <- numeric()
  
  mu[1] <- rnorm(1, ybar, 100) 
  sigma2[1] <- rnorm(1, var(data), 10)
  
  for(i in 2:M) { 
    astar <- a + 0.5*n
    bstar <- b + 0.5*sum((data - mu[i-1])^2)
    sigma2[i] <- rinvgamma(1, shape=astar, rate=bstar)
    
    mstar <- (n*ybar/sigma2[i] + m/v)/(n/sigma2[i] + 1/v)
    vstar <- 1/(n/sigma2[i] + 1/v)
    mu[i] <- rnorm(1, mstar, sqrt(vstar))
  }
  
  out <- cbind(mu, sigma2)[-c(1:burn),]
  colnames(out) <- c("mu","sigma2")
  out
}


# run the gibbs sampler for the 60s prior to get 60s posterior
# we used m=180 and v=400 to describe our uncertainty about mu, and a=2 and 
# b=3600 to desribe our uncertainty about the variance sigma2
post_60s <- unit8.gibbs(duration_60, m=180, v=400, a=2, b=3600, M=10000, burn=1000)

# check the convergence for mu and sigma2
plot(post_60s[,1], type="l")
plot(post_60s[,2], type="l")

# check the mixing with acf plots
acf(post_60s[,1])
acf(post_60s[,2])

# 95 % credible interval on the posterior mu
quantile(post_60s[,1], c(0.025, 0.975))


# plot the posterior draws for the mean (post_60s[,1])
plot(seq(140,220,length=1001), dnorm(seq(100,260,length=1001), 
          m_60, sqrt(v_60)), type = "l", main="Prior vs. Posterior for 60s song 
     length", xlab= "Song Length in Seconds", ylab="Density", ylim=c(0,.075), 
     col="red");
legend(x="topleft", legend=c("Prior","Posterior", "Credible Interval Bounds"), 
       col=c("red", "blue", "darkgreen"), lty=1)
lines(density(post_60s[,1]), col="blue")
abline(v=quantile(post_60s[,1], c(0.025, 0.975)), col="darkgreen", lty=2)

# Now lets look at post_60s[,2], sigma2
mean(post_60s[,2])
plot(density(post_60s[,2]))

# find 95% credible interval for variance
quantile(post_60s[,2], c(0.025, 0.975))

# compare prior and posterior for variance
plot(seq(0,10000,length=1001), dinvgamma(seq(0,10000,length=1001), 2, 3600), 
     type="l", col = "red", ylim=c(0,0.0008), main="Prior vs. Posterior on 
     Variance", xlab="Variance of Song Length", ylab="Density");
legend(x="topright", legend=c("Prior", "Posterior", "Credible Interval"), col=
         c("red", "blue", "darkgreen"), lty=1)
lines(density(post_60s[,2]), col="blue")
abline(v=quantile(post_60s[,2], c(0.025, 0.975)), col="darkgreen", lty=2)
```



Next, lets do almost the same exact analysis, but for the song lengths of 
today's top songs
```{r}
# Prior parameter values
# Let's once again choose m=180, expecting songs to also be 3 minutes long today
# We would once again expect songs to vary by a minute on either end, 
# so lets have a variance v of (120/6)^2 = 400
m_today <- 180; v_today <- 400
# and again, lets choose the same values of a and b as before, having no real
# information that popular songs have differed all that much since the 60s
a <- 2; b <- 3600

# run the gibbs sampler for the todays prior to get todays posterior
# we used m=180 and v=400 to describe our uncertainty about mu, and a=2 and 
# b=3600 to desribe our uncertainty about the variance sigma2
post_today <- unit8.gibbs(duration_today, m=180, v=400, a=2, b=3600, M=10000, 
                          burn=1000)

# check the convergence for mu and sigma2
plot(post_today[,1], type="l")
plot(post_today[,2], type="l")

# check the mixing with acf plots
acf(post_today[,1])
acf(post_today[,2])

# 95 % credible interval on the posterior mu
quantile(post_today[,1], c(0.025, 0.975))


# plot the posterior draws for the mean (post_60s[,1])
plot(seq(140,220,length=1001), dnorm(seq(100,260,length=1001), 
          m_60, sqrt(v_60)), type = "l", main="Prior vs. Posterior for today's 
          song length", xlab= "Song Length in Seconds", ylab="Density", 
     ylim=c(0,.075), col="red");
legend(x="topleft", legend=c("Prior","Posterior", "Credible Interval Bounds"), 
       col=c("red", "blue", "darkgreen"), lty=1)
lines(density(post_today[,1]), col="blue")
abline(v=quantile(post_today[,1], c(0.025, 0.975)), col="darkgreen", lty=2)

# Now lets look at post_60s[,2], sigma2

# find 95% credible interval for variance
quantile(post_today[,2], c(0.025, 0.975))

# compare prior and posterior for variance
plot(seq(0,10000,length=1001), dinvgamma(seq(0,10000,length=1001), 2, 3600), 
     type="l", col = "red", ylim=c(0,0.0013), main="Prior vs. Posterior on 
     Variance", xlab="Variance of Song Length", ylab="Density");
legend(x="topright", legend=c("Prior", "Posterior", "Credible Interval"), col=
         c("red", "blue", "darkgreen"), lty=1)
lines(density(post_today[,2]), col="blue")
abline(v=quantile(post_today[,2], c(0.025, 0.975)), col="darkgreen", lty=2)

```


Now, let's analyze the difference between our posterior draws from the 60s vs
today's top songs.

```{r}
# Create the vectors containing the difference for mu and sigma2
diff_mu <- post_60s[,1] - post_today[,1]
diff_sigma2 <- post_60s[,2] - post_today[,2]

# find 95% credible interval on the difference
quantile(diff_mu, c(0.025, 0.975))

plot(density(diff_mu), main="Difference in means; 60s - Today", 
     xlab= "Difference in mean mu")
abline(v=quantile(diff_mu, c(0.025, 0.975)), col="darkgreen", lty=2)

# Now for the difference in sigma2, variance
# lets find the 95% credible interval first
quantile(diff_sigma2, c(0.025, 0.975))

plot(density(diff_sigma2), main="Difference in variance; 60s - Today", 
     xlab= "Difference in variance sigma2")
abline(v=quantile(diff_sigma2, c(0.025, 0.975)), col="darkgreen", lty=2)

```

